---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```
```

# Task 1: Process and summarize your data 

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
Research Question 1: How does the average “radius_mean” differ between
malignant (M) and benign (B) diagnoses, and is the difference
statistically significant?

Research Question 2: Is there a relationship between “perimeter_mean”
and “area_mean” in the dataset, and does this relationship differ
between malignant (M) and benign (B) diagnoses?

Research Question 3: Do specific combinations of features (e.g.,
“radius_mean,” “texture_mean,” “smoothness_mean”) serve as better
predictors for malignancy than individual features alone?

Research Question 4: How do the distributions of various “worst” features (e.g.,
“radius_worst,” “texture_worst”) differ between malignant and benign
diagnoses, and can these features help assess cancer severity?
<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

#Research Question 1: How does the average “radius_mean” differ between malignant (M) and benign (B) diagnoses, and is the difference statistically significant?

#Summarizing : Compute the range, mean, and two other summary statistics of "radius_mean" across the groups of the "diagnosis" variable.
# Compute summary statistics for "radius_mean"
```{r}

radius_mean_summary <- cancer_sample %>%
  group_by(diagnosis) %>%
  summarize(
    Range = diff(range(radius_mean)),
    Mean = mean(radius_mean),
    Median = median(radius_mean),
    SD = sd(radius_mean)
  )
print(radius_mean_summary)
```
The results show the summary statistics for "radius_mean" for both malignant (M) and benign (B) diagnoses.

#Graphing : Make a graph where it makes sense to customize the alpha transparency.

```{r}
# Create a box plot for "radius_mean" by diagnosis with customized alpha transparency
ggplot(cancer_sample, aes(x = diagnosis, y = radius_mean, fill = diagnosis)) +
  geom_boxplot(alpha = 0.5) +
  labs(x = "Diagnosis", y = "Radius Mean") +
  scale_fill_manual(values = c("M" = "red", "B" = "blue")) +
  ggtitle("Differences in Radius Mean by Diagnosis") +
  theme_minimal()

```
This code generates a box plot that visualizes the differences in "radius_mean" between malignant and benign diagnoses.

#Research Question 2: Is there a relationship between "perimeter_mean" and "area_mean" in the dataset, and does this relationship differ between malignant (M) and benign (B) diagnoses?

Summarizing: Compute the range, mean, and two other summary statistics of "radius_mean" across the groups of the "diagnosis" variable.


```{r}
# Compute the range, mean, and other summary statistics for "perimeter_mean" and "area_mean"
perimeter_mean_area_mean_summary <- cancer_sample %>%
  group_by(diagnosis) %>%
  summarize(
    Perimeter_Mean_Range = max(perimeter_mean) - min(perimeter_mean),
    Perimeter_Mean_Mean = mean(perimeter_mean),
    Perimeter_Mean_SD = sd(perimeter_mean),
    Area_Mean_Range = max(area_mean) - min(area_mean),
    Area_Mean_Mean = mean(area_mean),
    Area_Mean_SD = sd(area_mean)
  )

print(perimeter_mean_area_mean_summary)

```
In this code, we calculate summary statistics for "perimeter_mean" and "area_mean" separately for malignant (M) and benign (B) diagnoses. This helps us understand the central tendency and variability of these variables within each diagnosis group.

#Graphing:Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.

```{r}
# Create a scatter plot of "perimeter_mean" vs. "area_mean" with a logarithmic scale
ggplot(cancer_sample, aes(x = log(area_mean), y = perimeter_mean, color = diagnosis)) +
  geom_point() +
  labs(x = "Log(Area Mean)", y = "Perimeter Mean", color = "Diagnosis") +
  ggtitle("Relationship Between Perimeter Mean and Log(Area Mean)") +
  theme_minimal()

```
In this graph, we use a logarithmic scale for the "area_mean" variable to better visualize the relationship with "perimeter_mean." Differentiating points by diagnosis allows us to see if there are any patterns or differences in the relationship between these variables for malignant and benign cases.

#Research Question 3: Do specific combinations of features (e.g., "radius_mean," "texture_mean," "smoothness_mean") serve as better predictors for malignancy than individual features alone?

#Summarizing: Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.


```{r}

# Create a new variable for the combination of features
cancer_sample <- cancer_sample %>%
  mutate(combined_feature = radius_mean * texture_mean * smoothness_mean)

# Compute summary statistics grouped by diagnosis
combined_summary <- cancer_sample %>%
  group_by(diagnosis) %>%
  summarize(
    range = diff(range(combined_feature)),
    mean = mean(combined_feature),
    sd = sd(combined_feature),
    min = min(combined_feature),
    max = max(combined_feature)
  )

# View the summary statistics
print(combined_summary)

```


#Graphing:Make a graph where it makes sense to customize the alpha transparency.

```{r}
# Create a scatterplot comparing combined_feature to radius_mean
scatterplot <- cancer_sample %>%
  ggplot(aes(x = combined_feature, y = radius_mean, color = diagnosis)) +
  geom_point(alpha = 0.5) +
  labs(x = "Combined Feature", y = "Radius Mean") +
  theme_minimal()

# Display the scatterplot
print(scatterplot)
```

In this scatterplot, we're comparing the combined feature on the x-axis and "radius_mean" on the y-axis, with different colors for malignant (M) and benign (B) diagnoses. This can help you visually assess whether the combined feature is a good predictor for malignancy in comparison to "radius_mean.


#Research Question 4: How do the distributions of various "worst" features (e.g., "radius_worst," "texture_worst") differ between malignant and benign diagnoses, and can these features help assess cancer severity?

#Summarizing:Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*

In this task, I create a categorical variable called "severity_category" based on the "radius_worst" variable to categorize it into different severity levels. I then count the frequency of severity categories by diagnosis and calculate the proportion of each category

```{r}
# Create a categorical variable "severity_category" for "radius_worst"
cancer_sample <- cancer_sample %>%
  mutate(severity_category = cut(radius_worst, breaks = c(0, 20, 25, 30, Inf),
                                 labels = c("Low", "Moderate", "High", "Very High")))

# Count the frequency of severity categories by diagnosis
severity_summary <- cancer_sample %>%
  group_by(diagnosis, severity_category) %>%
  summarize(Count = n()) %>%
  mutate(Proportion = Count / sum(Count))
print(severity_summary)
```
The code categorizes "radius_worst" into severity categories (Low, Moderate, High, and Very High) and then counts the frequency of each category for malignant (M) and benign (B) diagnoses. This summary helps assess how "radius_worst" is distributed across different severity levels for each diagnosis.


#Grapghing:Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.


```{r}
# Create three histograms with different-sized bins for "radius_worst"
ggplot(cancer_sample, aes(x = radius_worst, fill = diagnosis)) +
  geom_histogram(binwidth = 2, alpha = 0.6, position = "identity") +
  facet_wrap(~severity_category) +
  labs(x = "Radius Worst", y = "Count") +
  scale_fill_manual(values = c("M" = "red", "B" = "blue")) +
  ggtitle("Distribution of Radius Worst by Diagnosis and Severity Category") +
  theme_minimal()
```

This code generates histograms that display the distribution of "radius_worst" for different severity categories, helping visualize how this feature varies in terms of cancer severity. The use of facets and color differentiation by diagnosis enhances the understanding of these distributions.


### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

Research Question 1: The Summarizing task effectively addresses the research question by comparing key statistics, including range, mean, standard deviation, and median, for the "radius_mean" feature across malignant and benign tumors. The results clearly indicate that malignant tumors have a notably higher radius mean, aligning with the anticipated trend. To assess the statistical significance of this difference, further analysis such as a t-test may be necessary. 

Research Question 2:In the Summarizing task, I effectively addresses the research question by comparing key features, including range, mean, standard deviation, and median, for the perimeter_mean" and "area_mean" feature across malignant and benign tumors. The results clearly indicate that malignant tumors have a notably higher , aligning with the anticipated trend. In graphing I also compare the relationship between Perimeter_mean and Log(area_mean), it showed that Perimeter_mean and area mean are directly proportional to each.

Research Question 3: In this task I compared the combinations of features (e.g., "radius_mean," "texture_mean," "smoothness_mean") and made a scatter plot of combined features in relationship with the radius_mean and found that the radius_mean of malignant tumors have a notably higher relation with the combined features.

Research Question 4:In this task, I categorizes "radius_worst" into severity categories (Low, Moderate, High, and Very High) and then counts the frequency of each category for malignant (M) and benign (B) diagnoses and made histogram that display the distribution of "radius_worst" for different severity categories, helping visualize how this feature varies in terms of cancer severity and found out that the radius_worst of the benign tumor lies in low severity zone and most of the malignant radius_worst counts lies is moderate to high risk zone.  









# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

```{r}
cancer_subset <- cancer_sample %>%
  select(ID, diagnosis, radius_mean, texture_mean, perimeter_mean, area_mean, smoothness_mean, compactness_mean )
print(cancer_subset)
```

To ensure that each row represents a unique observation, I will examine the values in the ID column. If each ID is associated with a distinct patient without any duplicates, it confirms that each row indeed corresponds to a unique patient in the dataset.

```{r}
is_distinct <- n_distinct(cancer_subset$ID) == nrow(cancer_subset)

print(is_distinct)
```
To ensure that every cell in the dataset contains a valid value.
```{r}
cells_value <- cancer_subset%>% 
  summarise_all(~all(!is.na(.))) 

print(cells_value)
```
The initial code chunk involves the selection of 8 specific variables from the dataset. Subsequently, the second and third code chunks produce "TRUE" outcomes, serving as evidence that each row in the dataset represents a unique and distinct observation, and that every cell in all columns contains a valid value. This confirms that the data is indeed tidy.

### 2.2 (4 points)


Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

Before;
```{r}
cancer_sample
```

After,
#This code will pivot the selected subset into a long format where the "variable" column will contain the names of the original columns, and the "value" column will contain the corresponding values.
```{r}
untidy_cancer <- cancer_subset %>%
  pivot_longer(
    cols = -c(ID, diagnosis),
    names_to = "variable",
    values_to = "value"
  )

untidy_cancer
```

#Now tidy it back to the Original State.
This will restore the data to its original wide format.
```{r}
tidy_cancer <- untidy_cancer %>%
  pivot_wider(
    names_from = variable,
    values_from = value
  )

tidy_cancer
```


### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

1.  Does the difference between radius_mean between Malignant and Benign tumor is significantly differ.
2.  Which feature exhibit the largest difference in mean between malignant and benign diagnose?

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

For Question 1, We already found that there is a different in radius_mean in between malignant and benign tumor, but to be sure the difference is real we need to find out if the difference is significant or not. Aslo for Question 2, if we could find out which feature mean exhibit the largest difference, then it could serve as the best initial diagnosis.


Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 


1. Does the difference between "radius_mean" between Malignant and Benign tumors significantly differ?

For this question, we only need the "radius_mean" and "diagnosis" columns.

```{r}
# Select relevant columns
cancer_data_rq1 <- cancer_sample %>%
  select(diagnosis, radius_mean)

# Confirm that the data is tidy
head(cancer_data_rq1)
```
2. Which feature exhibits the largest difference in mean between malignant and benign diagnoses?

For this question, we need all numerical features except "ID."

```{r}
# Select relevant columns (all numerical features except ID)
cancer_data_rq2 <- cancer_sample %>%
  select(-ID, where(is.numeric))

# Confirm that the data is tidy
head(cancer_data_rq2)
```

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.



**Research Question**: Does the difference between "radius_mean" between Malignant and Benign tumors significantly differ?

**Variable of interest**: radius_mean

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

```{r}

t_test_result <- t.test(cancer_data_rq1$radius_mean ~ cancer_data_rq1$diagnosis)

t_test_result
```

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.


I will produce a relevant p-value from the t-test I performed earlier to assess whether there is a significant difference in the mean "radius_mean" between malignant and benign tumors.

I will use the broom package to extract the p-value:

```{r}

library(broom)

p_value <- tidy(t_test_result)$p.value

p_value


```
The p-value for the t-test comparing the mean "radius_mean" between malignant and benign tumors is approximately 1.68 x 10^(-64), which is an extremely small value. This indicates that the difference in "radius_mean" between malignant and benign tumors is highly statistically significant.

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

```{r}
output_dir <- "C:\Users\user\Downloads"

write_csv(radius_mean_summary, here::here("output", "MDA2_radius_mean_summary.csv"))
```



## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.


```{r}

saveRDS(t_test_result, here::here("output", "t_test_result.RDS"))
#Save the t_test_result in a file named “t_test_result.RDS” to a folder named “output”
readRDS(here::here("output", "t_test_result.RDS"))

```
# Save the model object to an RDS file
model <- lm(radius_mean ~ diagnosis, data = cancer_data_rq1)
saveRDS(model, file = here::here("output", "model_radius_mean.rds"))

# Load the model object from the RDS file
loaded_model <- readRDS(here::here("output", "model_radius_mean.rds"))

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

### Attribution

Thanks to Victor Yuan for mostly putting this together.

